---
title: "Nursing Home Project Tree Bagging Random Forest 2"
author: "Silvia"
date: '2023-01-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tree, Bagging, Random Forest \
Some pre-processing steps on the dataset:

```{r}
#import and rename the dataset
datafile <- read.csv("WiscNursingHome.csv")
#Eliminate the missing values
datafile <- na.omit(datafile)
#Factorization of the categorical variables
datafile$ORGSTR <- as.factor(datafile$ORGSTR)
datafile$URBAN <- as.factor(datafile$URBAN)
datafile$MSA <- as.factor(datafile$MSA)
datafile$PRO <- as.factor(datafile$PRO)
datafile$TAXEXEMPT <- as.factor(datafile$TAXEXEMPT)
datafile$SELFFUNDINS <- as.factor(datafile$SELFFUNDINS)
datafile$MCERT <- as.factor(datafile$MCERT)
```

Some libraries: \ 

```{r}
library("rpart") #tree
library("rpart.plot") #tree plot
library("Metrics")  # rmse
library("ipred") #Bagging, first method
library("caret") #Bagging, second method
library("randomForest") # Random Forest
library("plyr")
```

## Regression trees: \
### Random split: training set(70%) and test set(30%) \

```{r}
set.seed(123)

sample <- sample(c(TRUE, FALSE), nrow(datafile), replace=TRUE, prob=c(0.7,0.3))
train <- datafile[sample, ]
test  <- datafile[!sample, ]
```


### Maximal complexity (cp=0) - Random split \

```{r}
set.seed(123)
fit_tree <- rpart(TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR, data=train, method = "anova", control = rpart.control(cp = 0))
printcp(fit_tree) #cp table
plotcp(fit_tree) #plot cp values, error, splits
plot(fit_tree, uniform = TRUE, main = "Regression Tree, cp=0") #Tree plot
text(fit_tree, use.n = TRUE, all = TRUE, cex = .8)
pred <- predict(fit_tree, newdata = test)
rmse(test$TPY, pred)
SSE <- sum((pred - test$TPY)^2)
SST <- sum((test$TPY - mean(test$TPY))^2)
R_square <- 1 - SSE / SST
R_square
```


### Maximal complexity (cp=0) - 10 fold CV \

```{r}
set.seed(123)
form <- "TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR"
folds <- split(datafile, cut(sample(1:nrow(datafile)),10))
err <- rep(NA, length(folds))
Rsq <- rep(NA, length(folds))

for (i in 1:length(folds)) {
  test <- ldply(folds[i], data.frame)
  train <- ldply(folds[-i], data.frame)
  tmp.model <- rpart(form , train, method = "anova", control = rpart.control(cp = 0))
  tmp.predict <- predict(tmp.model, newdata = test)
  err[i] <- rmse(test$TPY, tmp.predict)
  Rsq[i] <- 1 - (sum((tmp.predict - test$TPY)^2))/ (sum((test$TPY - mean(test$TPY))^2))
}
print(sprintf("average rmse using k-fold cross-validation: %.3f", mean(err)))
print(sprintf("average R squared using k-fold cross-validation: %.3f", mean(Rsq)))
```

### Automatic choice of the cp (cp=) - Random split \

```{r}
set.seed(123)
fit_tree <- rpart(TPY ~ SQRFOOT + MSA + SELFFUNDINS + MCERT + ORGSTR, data=train, method = "anova")
printcp(fit_tree) #cp table
plotcp(fit_tree) #plot cp values, error, splits
plot(fit_tree, uniform = TRUE, main = "Regression Tree, cp=automatic") #Tree plot
text(fit_tree, use.n = TRUE, all = TRUE, cex = .8)
pred <- predict(fit_tree, newdata = test)
rmse
SSE <- sum((pred - test$TPY)^2)
SST <- sum((test$TPY - mean(test$TPY))^2)
R_square <- 1 - SSE / SST
R_square
```

### Automatic choice of the cp (cp=) - 10 fold CV \

```{r}
set.seed(123)
form <- "TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR"
folds <- split(datafile, cut(sample(1:nrow(datafile)),10))
err <- rep(NA, length(folds))
Rsq <- rep(NA, length(folds))

for (i in 1:length(folds)) {
  test <- ldply(folds[i], data.frame)
  train <- ldply(folds[-i], data.frame)
  tmp.model <- rpart(form , train, method = "anova")
  tmp.predict <- predict(tmp.model, newdata = test)
  err[i] <- rmse(test$TPY, tmp.predict)
  Rsq[i] <- 1 - (sum((tmp.predict - test$TPY)^2))/ (sum((test$TPY - mean(test$TPY))^2))
}
print(sprintf("average rmse using k-fold cross-validation: %.3f", mean(err)))
print(sprintf("average R squared using k-fold cross-validation: %.3f", mean(Rsq)))
```

### Manual set of the cp (cp=0.0017) - Random split \

```{r}
set.seed(123)
fit_tree <- rpart(TPY ~ SQRFOOT + MSA + SELFFUNDINS + MCERT + ORGSTR, data=train, method = "anova", control = rpart.control(cp = 0.0017))
printcp(fit_tree) #cp table
plotcp(fit_tree) #plot cp values, error, splits
plot(fit_tree, uniform = TRUE, main = "Regression Tree, cp=0.0017") #Tree plot
text(fit_tree, use.n = TRUE, all = TRUE, cex = .8)
pred <- predict(fit_tree, newdata = test)
rmse(test$TPY, pred)
SSE <- sum((pred - test$TPY)^2)
SST <- sum((test$TPY - mean(test$TPY))^2)
R_square <- 1 - SSE / SST
R_square
```

### Manual set of the cp (cp=0.0017) - 10 fold CV \

```{r}
set.seed(123)
form <- "TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR"
folds <- split(datafile, cut(sample(1:nrow(datafile)),10))
err <- rep(NA, length(folds))
Rsq <- rep(NA, length(folds))

for (i in 1:length(folds)) {
  test <- ldply(folds[i], data.frame)
  train <- ldply(folds[-i], data.frame)
  tmp.model <- rpart(form , train, method = "anova", control = rpart.control(cp = 0.0017))
  tmp.predict <- predict(tmp.model, newdata = test)
  err[i] <- rmse(test$TPY, tmp.predict)
  Rsq[i] <- 1 - (sum((tmp.predict - test$TPY)^2))/ (sum((test$TPY - mean(test$TPY))^2))
}
print(sprintf("average rmse using k-fold cross-validation: %.3f", mean(err)))
print(sprintf("average R squared using k-fold cross-validation: %.3f", mean(Rsq)))
```

## Bagging \
1)ipred package \
First, let find the best number of trees to use: \

```{r}
ntree <- seq(10, 50, 2)
rmse <- vector(mode = "numeric", length = length(ntree))
for (i in seq_along(ntree)) {
  set.seed(123)
  fit_bag <- bagging(TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR, data=train, coob = TRUE, nbagg =        ntree[i])
  rmse[i] <-fit_bag$err
}

plot(ntree, rmse, type = 'l', lwd = 2)
abline(v = 44, col = "red", lty = "dashed")
```

For the dataset the best number of trees is 44. \

Bagging with 44 trees, random split: \

```{r}
set.seed(123)

# train 
fit_bag16 <- bagging(TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR, nbagg=44, data=train, coob = TRUE)

fit_bag16

pred <- predict(fit_bag16, test)
rmse(pred, test$TPY)
SSE <- sum((pred - test$TPY)^2)
SST <- sum((test$TPY - mean(test$TPY))^2)
R_square <- 1 - SSE / SST
R_square
```

Bagging with 44 trees, 10 fold CV: \

```{r}
set.seed(123)
folds <- split(datafile, cut(sample(1:nrow(datafile)),10))
err <- rep(NA, length(folds))
Rsq <- rep(NA, length(folds))

for (i in 1:length(folds)) {
  test <- ldply(folds[i], data.frame)
  train <- ldply(folds[-i], data.frame)
  tmp.model <- bagging(TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR, nbagg=44, data=train, coob = TRUE)
  tmp.predict <- predict(tmp.model, newdata = test)
  err[i] <- rmse(test$TPY, tmp.predict)
  Rsq[i] <- 1 - (sum((tmp.predict - test$TPY)^2))/ (sum((test$TPY - mean(test$TPY))^2))
}
print(sprintf("average rmse using k-fold cross-validation: %.3f", mean(err)))
print(sprintf("average R squared using k-fold cross-validation: %.3f", mean(Rsq)))
```

2)caret package (non capisco se fa già cross validation da solo, e se si, quale dataset usare per predict)\

```{r}
# Specify 10-fold cross validation
set.seed(123)
ctrl <- trainControl(method = "cv",  number = 10) 

# CV bagged model
fit_bag <- train(TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR, data = train, method = "treebag", trControl = ctrl, importance = TRUE)

fit_bag

plot(varImp(fit_bag), 20) 

pred <- predict(fit_bag, test)
rmse(pred, test$TPY)
SSE <- sum((pred - test$TPY)^2)
SST <- sum((test$TPY - mean(test$TPY))^2)
R_square <- 1 - SSE / SST
R_square
```

## Random Forest \
Random forest with 500 trees(default, but good) \
Random split: \

```{r}
set.seed(123)
fit_forest <- randomForest(TPY ~ SQRFOOT + URBAN + PRO + TAXEXEMPT + MSA + SELFFUNDINS + MCERT + ORGSTR, data = train, proximity=TRUE) 
plot(fit_forest)

pred <- predict(fit_forest, test)
rmse(pred, test$TPY)
SSE <- sum((pred - test$TPY)^2)
SST <- sum((test$TPY - mean(test$TPY))^2)
R_square <- 1 - SSE / SST
R_square
```

10 fold CV

```{r}
set.seed(123)
folds <- split(datafile, cut(sample(1:nrow(datafile)),10))
err <- rep(NA, length(folds))
Rsq <- rep(NA, length(folds))

for (i in 1:length(folds)) {
  test <- ldply(folds[i], data.frame)
  train <- ldply(folds[-i], data.frame)
  tmp.model <- randomForest(TPY ~ SQRFOOT + MSA + URBAN + PRO + TAXEXEMPT + SELFFUNDINS + MCERT + ORGSTR, data=train, coob = TRUE)
  tmp.predict <- predict(tmp.model, newdata = test)
  err[i] <- rmse(test$TPY, tmp.predict)
  Rsq[i] <- 1 - (sum((tmp.predict - test$TPY)^2))/ (sum((test$TPY - mean(test$TPY))^2))
}
print(sprintf("average rmse using k-fold cross-validation: %.3f", mean(err)))
print(sprintf("average R squared using k-fold cross-validation: %.3f", mean(Rsq)))
```

