---
title: "Filtered linear models"
author: "tommaso tarchi"
date: "2023-01-18"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## importing the dataset and omitting the missing values
datafile <- read.csv("../WiscNursingHome.csv")
datafile <- na.omit(datafile)

## dividing data from different years
data_2000 <- datafile[datafile$CRYEAR == 2000, ]
data_2001 <- datafile[datafile$CRYEAR == 2001, ]

datafile$ORGSTR <- as.factor(datafile$ORGSTR)
datafile$MSA <- as.factor(datafile$MSA)
```

We noticed that the thing that all the outliers had in common was the high rate of utilization TPY/NUMBED. Therefore we can compute the new variable UTILIZATION_RATE and build a new dataset excluding from the observation of the year 2000 the first and the last quartiles w.r.t. to that variable (i.e. the outliers):
```{r}
data_2000$UTILIZATION_RATE <- data_2000$TPY / data_2000$NUMBED
summary(data_2000$UTILIZATION_RATE)

data_trunc <- data_2000[data_2000$UTILIZATION_RATE > 0.89 & data_2000$UTILIZATION_RATE < 0.96, ]
```


## Quatitative variables only

```{r}
summary(lm(TPY ~ NUMBED + SQRFOOT, data = data_trunc))
```
```{r}
par(mfrow = c(2, 2))
plot(lm(TPY ~ NUMBED + SQRFOOT, data = data_trunc))
```

Let's see if we get better residuals with log-transformed variables:
```{r}
summary(lm(log(TPY) ~ log(NUMBED) + log(SQRFOOT), data = data_trunc))
```

```{r}
par(mfrow = c(2, 2))
plot(lm(log(TPY) ~ log(NUMBED) + log(SQRFOOT), data = data_trunc))
```

Still the results are not satisfactory: we managed to fix homoscedasticity but we could not get normally distributed residuals.


## Adding categorical variables

```{r}
summary(lm(log(TPY) ~ log(NUMBED) + TAXEXEMPT, data = data_trunc))
```
```{r}
anova(glm(log(TPY) ~ log(NUMBED) + TAXEXEMPT, data = data_trunc, family = gaussian),
      test = "Chisq")
```
```{r}
par(mfrow = c(2, 2))
plot(lm(log(TPY) ~ log(NUMBED) + TAXEXEMPT, data = data_trunc))
```