
---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

**Brief Introduction**

*The nursing home data are provided by the Wisconsin Department of Health and Family Services (DHFS). The State of Wisconsin Medicaid program funds nursing home care for individuals qualifying on the basis of need and financial status. As part of the conditions for participation, Medicaid-certified nursing homes must file an annual cost report to DHFS, summarizing the volume and cost of care provided to all of its residents, Medicaid funded and otherwise. These cost reports are audited by DHFS staff and form the basis for facility-specific Medicaid daily payment rates for subsequent periods. Note: A nursing home is a facility for the residential care of elderly or disabled people.*

**Objective**

*The aim of this project is to develop a predictive model that provide reliable utilization forecast to update their Medicaid funding rate schedule of nursing facilities.*

**Data**

*The data here is in cost report years 2000 and 2001. There are 362 facilities in 2000 and 355 facilities in 2001. Typically, utilization of nursing home care is measured in patient days (“patient days” is the number of days each patient was in the facility, summed over all patients).*



#### **Exploratory Data Analysis**



**Information about the variables**

| File Name:WiscNursingHome         | Number of obs: 717        | Number of variables: 12                       |
|:---------------------------------:|:-------------------------:|:---------------------------------------------:|
| **Variable (type of variable)**   | **Number of Obs Missing** | **Description**                               |
| hospID (categorical)              |                           | Hospital identification number                |
| CRYEAR (categorical binary)       |                           | Cost report year                              |
| TPY (quantitative continuous)     |                           | Total patient years                           |
| NUMBED (quantitative discrete)    |                           | Number of beds                                |
| SQRFOOT (quantitative continuous) |    10                     | Square footage of the nursing home            |
| MSA (categorical)                 |                           | Metropol Statist Area code, 1-13, 0 for rural |
| URBAN (categorical Binary)        |                           | 1 if urban, 0 if rural                        |
| PRO (categorical Binary)          |                           | 1 if for profit, 0 for non-profit             |
| TAXEXEMPT (categorical Binary)    |                           | 1 if tax-exempt                               |
| SELFFUNDINS (categorical Binary)  |                           | 1 if self-funded for insurance                |
| MCERT (categorical Binary)        |                           | 1 if Medicare certified                       |
| ORGSTR (categorical)              |                           | 1 for profit, 2 for tax-exempt,3 for gov unit |



```{r}
datafile <- read.csv("WiscNursingHome.csv")
datafile
```
**Analyzing variables distribution**
```{r}
# Set the layout of the graph
par(mfrow = c(3,4))

# Draw the first histogram
hist(datafile$TPY, main = "Histogram of TPY", xlab = "TPY", col = "blue")

# Draw the second histogram
hist(datafile$NUMBED, main = "Histogram of NUMBED", xlab = "NUMBED", col = "blue")

# Draw the third histogram
hist(datafile$SQRFOOT, main = "Histogram of SQRFOOT", xlab = "SQRFOOT", col = "blue")
hist(datafile$MSA, main = "Histogram of MSA", xlab = "MSA", col = "purple")
hist(datafile$URBAN, main = "Histogram of URBAN", xlab = "URBAN", col = "green")
hist(datafile$PRO, main = "Histogram of PRO", xlab = "PRO", col = "green")
hist(datafile$TAXEXEMPT, main = "Histogram of TAXEXEMPT", xlab = "TAXEXEMPT", col = "green")
hist(datafile$SELFFUNDINS, main = "Histogram of SELFFUNDINS", xlab = "SELFFUNDINS", col = "green")
hist(datafile$MCERT, main = "Histogram of MCERT", xlab = "MCERT", col = "green")
hist(datafile$ORGSTR, main = "Histogram of ORGSTR", xlab = "ORGSTR", col = "purple")
```

##### **Analysis of quantitative variables**


*Let's see the summary statistics*

```{r}
summary(datafile)
```

*For categorical variables we can observe that they are unbalance. On the other hand, for the quantitative variables, we can see the difference between the median and the mean which highlight the fact that the distribution is right skewed. Let's see this more clearly the distribution for the quantitative variables through a Box plot chart. *

```{r}

par(mfrow = c(1, 3))

boxplot(datafile$TPY, main = "Boxplot TPY")
 
boxplot(datafile$NUMBED, main = "Boxplot NUMBED")

boxplot(datafile$SQRFOOT, main = "Boxplot SQRFOOT")
```



*We can see clearly that the distributions are right skewed, there are many outliers (observations outside of the whisker). In addtion, the means of the TRY and NUMBED variables are close, while the mean for SRQFOOT is less than the half the others.*


```{r}
library(ggplot2)
options(warn = - 1)
# install.packages("GGally")
library(GGally)
ggpairs(subset(datafile, select = c(TPY, SQRFOOT, NUMBED)))
```

*From the histograms and pair plots for the the non binary variables, we can see that the quantitative variables are skewed to the right. Then a log normal transformation of the variables should be useful for the statistical analysis*  
*The variable MSA have a bi modal distribution with modes in approximately 0 and 12.*
*With respect to the correlation, the quantitative variables TRY and NUMBED are highly correlated (0,984). This high correlation means that both variables have the same information, which should be redundant to include in a single model. On the other hand, the correlation of SQRFOOT with TRY and NUMBED is around 0.818. Let's see the normality of the quantitative variables distribution*

```{r}
par(mfrow = c(1,3))

qqnorm(datafile$TPY, main = 'Q-Q Plot for Normality for TPY', xlab = 'Theoretical Dist',
       ylab = 'Sample dist', col = 'steelblue')
qqline(datafile$TPY, col = 'red', lwd = 2, lty = 2)

qqnorm(datafile$NUMBED, main = 'Q-Q Plot for Normality for NUMBEB', xlab = 'Theoretical Dist',
       ylab = 'Sample dist', col = 'steelblue')
qqline(datafile$NUMBED, col = 'red', lwd = 2, lty = 2)

qqnorm(datafile$SQRFOOT, main = 'Q-Q Plot for Normality for SQRFOOT', xlab = 'Theoretical Dist',
       ylab = 'Sample dist', col = 'steelblue')
qqline(datafile$SQRFOOT, col = 'red', lwd = 2, lty = 2)
```

*As we can observe in the Q-Q plots none of the quantitative variables has a normal distribution* 

*Our aim is to predict the utilization the nursing houses and the candidate response variable is TPY. Let's see the values for the mean and variance:*

```{r}
mean(datafile$TPY)
```

```{r}
var(datafile$TPY)
```

*Since this variable is right skewed distributed, continuous, and with a mean and variance significantly different, we assume that the variable TPY follows a Gamma distribution. Let's shows this through a QQ plot chart*

```{r}
#CHECK WHETHER OR NOT TPY IS GAMMA DISTRIBUTED
library(MASS)
library(qqplotr)
TPY_quantiles <- datafile$TPY[datafile$TPY < 250]
fit <- fitdistr(datafile$TPY, "gamma")
qgamma <- rgamma(701, shape=fit$estimate[1], rate = fit$estimate[2])
qqplot(TPY_quantiles, qgamma, main = "Gamma Q-Q Plot for TPY")
abline(0,1, col="red")
```

*The qqplot chart shows that our hypothesis about the distribution of the variable TPY is correct since the observations follow a gamma distribution*  


**Let's transform the skewed quantitative variables with the log function**

```{r}
library(dplyr)
datafile_log <- datafile %>% mutate(TPY_log = log(TPY), NUMBED_log = log(NUMBED), SQRFOOT_log = log(SQRFOOT))
datafile_log
```

*Let's see the summary statistics for the log transformed variables*

```{r}
summary(subset(datafile_log, select = c(TPY_log, NUMBED_log, SQRFOOT_log, TPY, SQRFOOT, NUMBED)))
```

*It is clear that now the variables have a log normal distribution since the mean and the median are much closer as well as the first and fourth quantile, let's see this graphically with histograms a box plots*

```{r}
# Set the layout of the graph
par(mfrow = c(2,3))

# Draw the first histogram
hist(datafile_log$TPY, main = "Histogram of TPY", xlab = "TPY", col = "blue")

# Draw the second histogram
hist(datafile_log$NUMBED, main = "Histogram of NUMBED", xlab = "NUMBED", col = "blue")

# Draw the third histogram
hist(datafile_log$SQRFOOT, main = "Histogram of SQRFOOT", xlab = "SQRFOOT", col = "blue")
hist(datafile_log$TPY_log, main = "Histogram of TPY_log", xlab = "TPY_log", col = "orange")
hist(datafile_log$NUMBED_log, main = "Histogram of NUMBED_log", xlab = "NUMBED_log", col = "orange")
hist(datafile_log$SQRFOOT_log, main = "Histogram of SQRFOOT_log", xlab = "SQRFOOT_log", col = "orange")

```



```{r}

par(mfrow = c(1, 3))

boxplot(datafile_log$TPY_log, main = "Boxplot TPY_log")
 
boxplot(datafile_log$NUMBED_log, main = "Boxplot NUMBED_log")

boxplot(datafile_log$SQRFOOT_log, main = "Boxplot SQRFOOT_log")
```


*After the log transformation of the variables, seems that many of the observations were considered outliers before, now they are included in the log-normal distribution. Besides, from the whiskers is easy to recognize that the distribution is approximately normal. Ending, in general, after the transformation, the mean of the three variables is closer. Let's see what information we can extract from a pairplot of the transformed variables*  

```{r}
options(warn = - 1)
library(ggplot2)
library(GGally)
ggpairs(subset(datafile_log, select = c(TPY_log, NUMBED_log, SQRFOOT_log)))
```


*As we can see in the pair plot chart, the values for the correlation between the log transformed variables are approximately the same than the original ones. However, increase the linear association of SQRFOOT with variables TPY and NUMBED. Let's see the normality of the log transformed quantitative variables*

```{r}
par(mfrow = c(1,3))

qqnorm(datafile_log$TPY_log, main = 'Q-Q Plot for Normality for Log(TPY)', xlab = 'Theoretical Dist',
       ylab = 'Sample dist', col = 'steelblue')
qqline(datafile_log$TPY_log, col = 'red', lwd = 2, lty = 2)

qqnorm(datafile_log$NUMBED_log, main = 'Q-Q Plot for Normality for Log(NUMBED)', xlab = 'Theoretical Dist',
       ylab = 'Sample dist', col = 'steelblue')
qqline(datafile_log$NUMBED_log, col = 'red', lwd = 2, lty = 2)

qqnorm(datafile_log$SQRFOOT_log, main = 'Q-Q Plot for Normality for Log(SQRFOOT)', xlab = 'Theoretical Dist',
       ylab = 'Sample dist', col = 'steelblue')
qqline(datafile_log$SQRFOOT_log, col = 'red', lwd = 2, lty = 2)
```

##### *Analysis of categorical variables*

*Histograms of the non binary categorical variables to see the distribution of the levels*

```{r}
# Set the layout of the graph
par(mfrow = c(1,2))
hist(datafile$MSA, main = "Histogram of MSA", xlab = "MSA", col = "purple", xlim = c(0, 14), ylim = c(0, 400))
hist(datafile$ORGSTR, main = "Histogram of ORGSTR", xlab = "ORGSTR", col = "purple", , ylim = c(0, 400))
```


*As we can see in the histogram for MSA the observations are concentrated in the levels 0 (rural) and 13 (Metropol. Statist. Area code). For the variable ORGSTR the observations are concentrated in the levels 1 for profit and 2 for tax-exempt*


*Let's see the relation of the responce and categorical variables*


```{r}
par(mfrow=c(1,2))
boxplot(TPY ~ MSA, data = datafile, horizontal = FALSE, 
        xlab = "MSA", ylab = "TPY")
```

*The boxplot for the distribution of TPY in the MSA levels (most frequent from previous hist.) shows that for level 0(rural) there are many outliers between 150 and 250 years with mean around 80 years and the rest of the observations below 150 years. For the level 13 (metropolitan area) there an extreme value that turns the distribution right skewed, with a mean around 100 years and the rest of the observations below 220 years.*

*Distribution of TPY over the rest of categorical Variables*

```{r}
par(mfrow=c(2,2))
boxplot(TPY ~ SELFFUNDINS, data = datafile, horizontal = TRUE, 
        ylab = "SELFFUNDINS", xlab = "TPY")
boxplot(TPY ~ TAXEXEMPT, data = datafile, horizontal = TRUE, 
        ylab = "TAXEXEMPT", xlab = "TPY")
boxplot(TPY ~ PRO, data = datafile, horizontal = TRUE, 
        xlab = "TPY", ylab = "PRO")
boxplot(TPY ~ ORGSTR, data = datafile, horizontal = TRUE, 
        xlab = "TPY", ylab = "ORGSTR")
```

*From the boxplots, we can see that there are no significant differences in the TPY distribution over the categorical variables. In the case of ORGSTR seems to be a significant difference between levels 1 (for profit) and 2 for tax-exempt with respect to the level (3 for gov unit) but since level 3 has less than 100 observations (10% approx) we can not affirm that difference is really significative.*

```{r}
par(mfrow = c(1, 2))
barplot(prop.table(table(datafile$PRO, datafile$TAXEXEMPT), margin = 1),
        beside=TRUE, xlab="TAX EXEMPT",col=c("gray","black"))
legend("topright", c("Non Profit", "Profit"), cex = 0.95, bty = "n",fill=c("gray","black"))
barplot(prop.table(table(datafile$PRO, datafile$SELFFUNDINS), margin = 1),
        beside=TRUE, xlab="SELFUNDINS",col=c("gray","black"))
legend("topleft", c("Non Profit", "Profit"), cex = 0.95, bty = "n",fill=c("gray","black"))
```

##### *Data pre-processing*


**Missing values processing**


```{r}
library(dplyr)
#DATA CLEANING
datafile <- na.omit(datafile) #drop NAs
```


*select only rows for Medicaid-certified nurse houses*

```{r}

datafile$UTILIZATION_RATE <- datafile$TPY / datafile$NUMBED
```

#datafile <- datafile %>% filter(MCERT == 1) 

**Transform into categorical variables**

```{r}
#Transform into categorical variables
datafile$ORGSTR <- as.factor(datafile$ORGSTR)
datafile$MSA <- as.factor(datafile$MSA)
```


**Selection, description, and possibly comparison of the most suitable statistical models**

```{r}
df_2000 <- datafile[datafile$CRYEAR == "2000", ]
df_2000
```


```{r}
df_2001 <- datafile[datafile$CRYEAR == "2001", ]
df_2001
```

```{r}
mean(df_2000$TPY) 
```
